Highly Available Transactions: Virtues and Limitations\iftechreport\\(Extended Version)\fi

### 2013-10-01 writeup/conclusion.tex
Editing pass

The current state of database software offers uncomfortable and
unnecessary choices between availability and transactional semantics.
Through our analysis and experiments, we have demonstrated how goals
of high availability will remain a critical aspect of many future data
storage systems. Accordingly, we expose a broad design space of Highly
Available Transactions (HATs), which can offer the key benefits of
highly available distributed systems---``always on'' operation during
partitions and low-latency operations---while also providing a family
of transactional isolation levels and replica semantics that have been
adopted in practice.  In taxonomizing this design space, we show that
many transactional guarantees are achievable with high availability,
while a few, like preventing Lost Update and Write Skew, are not. This
offers a useful connection between storage semantics studied in the
database and distributed systems communities.

HATs raise several challenges for future study. In this paper, we have
highlighted standardized models, but we would like a more concise
description of which models are and are not achievable in HAT
systems. Similarly, while we have studied the analytical and
experiment behavior of several HAT models, there is substantial work
in further understanding the performance and design of systems within
the large set of HAT models. We also believe it is time to revisit
techniques such as Sagas~\cite{sagas} and other long-running
transaction support in order to aid HAT programming. Weakened failure
assumptions as in escrow or in the form of bounded network asynchrony
could enable richer HAT semantics at the cost of general-purpose
availability. Alternatively, there is a range of possible hybrid
systems providing strong semantics during partition-free periods and
weakened semantics during partitions. Based on our understanding of
what is desired in the field and newfound knowledge of what is
possible to achieve, we believe HATs provide a large and useful design
space for exploration.

>>>>

The current state of database software offers uncomfortable and
unnecessary choices between availability and transactional semantics.
Through our analysis and experiments, we have demonstrated how goals
of high availability will remain a critical aspect of many future data
storage systems. Accordingly, we expose a broad design space of Highly
Available Transactions (HATs), which can offer the key benefits of
highly available distributed systems---``always on'' operation during
partitions and low-latency operations---while also providing a family
of transactional isolation levels and replica semantics that have been
adopted in practice.  In taxonomizing this design space, we show that
many transactional guarantees are achievable with high availability,
while a few, like preventing Lost Update and Write Skew, are
not. Surprisingly, many of the default (and sometimes strongest)
semantics provided by today's traditional database systems are
achievable as HATs, hinting that distributed databases need not
compromise availability in order to serve many existing applications.

HATs raise several challenges for future study. In this paper, we have
largely focused on previously defined isolation and data consistency
models from the consistency from the database and distributed systems
communities. Their previous definitions and, in many cases, widespread
adoption hints at their utility to end-users. However, we believe
there is considerable work to be done to improve the programmability
of highly available systems. Isolation and data consistency are means
by which \textit{application-level} consistency is achieved but are
typically not end goals for end-user applications. Our results hint
that a mix of HAT and non-HAT semantics (the latter used sparingly) is
required for practical applications, but the decision to employ each
and the system architectures for a hybrid approach remain open
problems. While we have studied the analytical and experiment behavior
of several HAT models, there is substantial work in further
understanding the performance and design of systems within the large
set of HAT models. Weakened failure assumptions as in escrow or in the
form of bounded network asynchrony could enable richer HAT semantics
at the cost of general-purpose availability. Alternatively, there is a
range of possible solutions providing strong semantics during
partition-free periods and weakened semantics during partitions. Based
on our understanding of what is desired in the field and newfound
knowledge of what is possible to achieve, we believe HATs provide a
large and useful design space for exploration.

### 2013-10-01 writeup/evaluation.tex
Editing pass

\vspace{.5em}\noindent\textbf{Non-serializability} Most existing
distributed implementations of weak isolation are not highly
available. Lock-based mechanisms such as those in Gray's original
proposal~\cite{gray-isolation} do not degrade gracefully in the
presence of partial failures. (Note, however, that lock-based
protocols \textit{do} offer the benefit of recency guarantees.) While
multi-versioned storage systems allow for a variety of transactional
guarantees, few offer traditional weak isolation (e.g.,
non-``tentative update'' schemes) in this context.  Chan and Gray's
read-only transactions have item-cut isolation with causal consistency
and RA (session \textit{PL-2L}~\cite{adya}) but are unavailable in the
presence of coordinator failure and assume serializable update
transactions~\cite{readonly}; this is similar to read-only and
write-only transactions more recently proposed by
Eiger~\cite{eiger}.  Brantner's S3 database~\cite{kraska-s3}
and Bayou~\cite{sessionguarantees} can all provide variants of session
\textit{PL-2L} with high availability, but none provide this HAT
functionality without substantial modification. Swift~\cite{swift} and
bolt-on causal consistency~\cite{bolton} are closest to providing
maximal sticky HAT semantics. Accordingly, it is possible to implement
many guarantees weaker than serializability---including guarantees
that are achievable with high availability---and still not achieve
high availability.

@@

Thus far, we have largely ignored the question of when HAT semantics
are useful (or otherwise are too weak). As we showed in
Section~\ref{sec:hats}, the main cost of high availability and low
latency comes in the inability to prevent Lost Update, Write Skew, and
provide recency bounds. In this section, we attempt to understand when
these guarantees matter both abstractly and in a representative
transactional application based on the TPC-C benchmark~\cite{tpcc}.

\vspace{.5em}\noindent\textbf{Commutativity and Monotonicity} As
evidenced by the CALM Theorem~\cite{calm} and by Commutative and
Replicated Data Types~\cite{crdt}, if updates logically commute, then
they can often be safely performed in different orders at different
replicas. Accordingly, as long as all writes are delivered to all
replicas, then a system executing monotonic logic may not suffer from
application-level consistency anomalies as a result of Lost Update or
Write Skew anomalies. However, applications with non-monotone state
mutation will not, in general, be able to maintain application-level
consistency constraints with HATs alone---in particular, applications
requiring visibility bounds should opt for unavailability.

@@

TPC-C consists of five transactions, capturing the operation of a
wholesale warehouse, including sales, payments, and deliveries. Two
transactions---\textit{Order-Status} and \textit{Stock-Level}---are
read-only and can be executed safely with HATs. Clients may read stale
data, but this does not violate TPC-C requirements and clients will
read their writes if they are sticky-available. Another transaction
type, \textit{Payment}, updates running balances for warehouses,
districts, and customer records and provides an audit trail. The
transaction is monotonic---increment- and append-only---so all balance
increase operations commute, and RA allows the maintenance of
foreign-key integrity constraints (e.g., via \texttt{UPDATE/DELETE
  CASCADE}).

@@

\vspace{.5em}\noindent\textit{Integrity Constraints.} Throughout execution, TPC-C also requires the maintenance of several
integrity constraints. For example, Consistency Condition 1 (3.3.2.1)
requires that each warehouse's sales count must reflect the sum of its
subordinate sales districts. This integrity constraint spans two
tables but, given the ability to update rows in both tables atomically
via RA, can be easily maintained. Consistency Conditions 4 through 12
(3.3.2.4-12) can similarly be satisfied by applying updates atomically
across tables. Consistency Conditions 2 and 3 (3.3.2.2-3) concern
order ID assignment and are problematic. Finally, while TPC-C is not
subject to multi-key anomalies, we note that many TPC-E isolation
tests (i.e., simultaneously modifying a product description and its
thumbnail) are also achievable using HATs.

@@

\vspace{.5em}\noindent\textit{Implementation.} Our prototype database
is a partially replicated (hash-based, partitioned) key-value backed
by LevelDB and implemented in Java using Apache Thrift. It currently
supports eventual consistency (hereafter, \texttt{eventual};
last-writer-wins RU with standard all-to-all anti-entropy between
replicas) and the efficient HAT RA algorithm as sketched in
Section~\ref{sec:ta}. (hereafter, \texttt{RA}). It also supports
non-HAT operation whereby all operations for a given key are routed to
a (randomly) designated \texttt{master} replica for each key
(guaranteeing single-key linearizability, as in Gilbert and Lynch's
CAP Theorem proof~\cite{gilbert-cap} and in PNUTS~\cite{pnuts}'s
\texttt{read-latest} operation; hereafter, \texttt{master}) as well as
distributed two-phase locking. Servers are durable: they synchronously
write to LevelDB before responding to client requests, while new
writes in RA are sent to a write-ahead log, which is synchronously
flushed to disk.

@@

\vspace{.5em}\noindent\textit{Geo-replication.} We first deploy the
database prototype across an increasing number of
datacenters. Figure~\ref{fig:wan-exp}A shows that, when operating two
clusters within a single datacenter, mastering each data item results
in approximately half the throughput and double the latency of
\texttt{eventual}. This is because HAT models are able to utilize
replicas in both clusters instead of always contacting the (single)
master. \texttt{RC}---essentially \texttt{eventual} with
buffering---is almost identical to \texttt{eventual}, while
\texttt{RA}---which incurs two writes for every client-side write
(i.e., new writes are sent to the WAL then subsequently moved into
LevelDB once stable)---achieves ~75\% of the throughput. Latency
increases linearly with the number of YCSB clients due to contention
within LevelDB.

In contrast, when the two clusters are deployed across the continental
United States (Figure~\ref{fig:wan-exp}B), the average latency of
\texttt{master} increases to $300$ms (a $278$--$4257\%$ latency
increase; average $37$ms latency per operation). For the same number
of YCSB client threads, \texttt{master} has substantially lower
throughput than the HAT configurations. Increasing the number of YCSB
clients \textit{does} increase the throughput of \texttt{master}, but
our Thrift-based server-side connection processing did not gracefully
handle more than several thousand concurrent connections. In contrast,
across two datacenters, the performance of eventual, RC, and RA are
near-identical to a single-datacenter deployment.

When five clusters (as opposed to two, as before) are deployed across
the five EC2 datacenters with lowest communication cost
(Figure~\ref{fig:wan-exp}C), the trend continues: \texttt{master}
latency increases to nearly $800$ms per transaction. As an attempt at
reducing this overhead, we implemented and benchmarked a variant of
quorum-based replication as in Dynamo~\cite{dynamo}, where clients
sent requests to all replicas, which completed as soon as a majority
of servers responded (guaranteeing regular
semantics~\cite{herlihy-art}); this strategy (not pictured) did not
substantially improve performance due to the network topology and
because worst-case server load was unaffected. With five clusters,
\texttt{RA}'s relative throughput decreased: every YCSB \texttt{put}
operation resulted in four \texttt{put} operations on remote replicas
and, accordingly, the cost of anti-entropy increased (e.g., each
server processed four replicas' anti-entropy as opposed to one before,
reducing the opportunity for batching and decreasing available
resources for incoming client requests). This in turn increased
garbage collection activity and, more importantly, IOPS when compared
to \texttt{eventual} and \texttt{RC}, causing \texttt{RA} throughput
to peak at around half of \texttt{eventual}. With in-memory
persistence (i.e., no LevelDB or WAL), \texttt{RA} throughput was
within 20\% of \texttt{eventual}.

@@

\begin{figure}[t!]
\begin{center}
\hspace{2em}\includegraphics[width=.8\columnwidth]{figs/strategylegend.pdf}\vspace{-2em}
\end{center}
\begin{center}
\includegraphics[width=\figfactor\columnwidth]{figs/finals/txnlen-thru.pdf}
\end{center}\vspace{-2.25em}
\caption{Transaction length versus throughput.}\vspace{-1em}
\label{fig:txlen}
\begin{center}
\includegraphics[width=\figfactor\columnwidth]{figs/finals/wprop-thru.pdf}
\end{center}\vspace{-2.25em}
\caption{Proportion of reads and writes versus throughput.}\vspace{-1em}
\label{fig:rprop}
\begin{center}
\includegraphics[width=\figfactor\columnwidth]{figs/finals/scaleout-thru.pdf}
%\includegraphics[width=\figfactor\columnwidth]{figs/finals/scaleout-thru-perserver.pdf}
\end{center}\vspace{-2.25em}
\caption{Scale-out of RA, Eventual, and RC.}\vspace{-1.5em}
\label{fig:scaleout}
\end{figure}

\vspace{.5em}\noindent\textit{Transaction length.} As shown in
Figure~\ref{fig:txlen} (clusters in Virginia and Oregon), throughput
of \texttt{eventual}, \texttt{RC}, and \texttt{master} operation are
unaffected by transaction length. In contrast, \texttt{RA} throughput
decreases linearly with increased transaction length: with $1$
operation per transaction, \texttt{RA} throughput is within 18\% of
\texttt{eventual} ($34$ bytes overhead), and with $128$ operations per
transaction, \texttt{RA} throughput is within $60\%$ ($1898$ bytes
overhead). This reflects our \texttt{RA} algorithm's metadata requirements,
which are proportional to transaction length and consume IOPS and
network bandwidth.

\vspace{.5em}\noindent\textit{Read proportion.} Our default (equal)
proportion of reads and writes is fairly pessimistic: for example,
Facebook reports $99.8\%$ reads for their workload~\cite{eiger}. As
shown in Figure~\ref{fig:rprop} (clusters in Virginia and Oregon),
with all reads, \texttt{RA} is within $4.8\%$ of \texttt{eventual};
with all writes, \texttt{RA} is within $33\%$, and the throughput of
\texttt{eventual} decreases by $288.8\%$ compared to all reads. At
$99.8\%$ reads, \texttt{RA} incurs a $7\%$ overhead ($5.8\%$ for in-memory
storage).

\vspace{.5em}\noindent\textit{Scale-out.} One of the key benefits of
our HAT algorithms is that they are shared-nothing, meaning they
should not compromise scalability. Figure~\ref{fig:scaleout} shows
that varying the number of servers across two clusters in Virginia and
Oregon (with $15$ YCSB clients per server) results in linear scale-out
for \texttt{eventual}, \texttt{RC}, and \texttt{RA}. \texttt{RC} and
\texttt{eventual} scale perfectly with an approximately $5$x
throughput increase moving from $5$ to $25$ servers per cluster. For
the same configuration, \texttt{RA} scales by $3.8$x, achieving over
$260,000$ operations per second. \texttt{RA} performance is due to
resource contention---with a memory-backed database (instead of
LevelDB), \texttt{RA} scales by $4.25$x (not shown)---and RA-related
performance heterogeneity across servers (Calvin's authors report
similar heterogeneity on EC2~\cite{calvin}). Nevertheless, this linear
scalability is not typical of traditional database system
designs~\cite{gray-isolation}.

\vspace{.5em}\noindent\textit{Summary.} Our experimental prototype
confirms our earlier analytical intuitions. HAT systems can provide
useful semantics without substantial performance penalties. In
particular, our RA algorithm can achieve throughput competitive with
eventual consistency at the expense of increased disk and network
utilization. Perhaps more importantly, all HAT algorithms circumvent
high WAN latencies inevitable with non-HAT implementations. Our results
highlight Deutsch's observation that ignoring factors such as latency
can ``cause big trouble and painful learning
experiences''~\cite{fallacies-deutsch}---in a single-site context,
paying the cost of coordination may be tenable, but, especially as
services are geo-replicated, costs increase.

>>>>

\vspace{.5em}\noindent\textbf{Non-serializability} Most existing
distributed implementations of weak isolation are not highly
available. Lock-based mechanisms such as those in Gray's original
proposal~\cite{gray-isolation} do not degrade gracefully in the
presence of partial failures. (Note, however, that lock-based
protocols \textit{do} offer the benefit of recency guarantees.) While
multi-versioned storage systems allow for a variety of transactional
guarantees, few offer traditional weak isolation (e.g.,
non-``tentative update'' schemes) in this context.  Chan and Gray's
read-only transactions have item-cut isolation with causal consistency
and MAV (session \textit{PL-2L}~\cite{adya}) but are unavailable in the
presence of coordinator failure and assume serializable update
transactions~\cite{readonly}; this is similar to read-only and
write-only transactions more recently proposed by
Eiger~\cite{eiger}.  Brantner's S3 database~\cite{kraska-s3}
and Bayou~\cite{sessionguarantees} can all provide variants of session
\textit{PL-2L} with high availability, but none provide this HAT
functionality without substantial modification. Swift~\cite{swift} and
bolt-on causal consistency~\cite{bolton} are closest to providing
maximal sticky HAT semantics. Accordingly, it is possible to implement
many guarantees weaker than serializability---including guarantees
that are achievable with high availability---and still not achieve
high availability.

@@

Thus far, we have largely ignored the question of when HAT semantics
are useful (or otherwise are too weak). As we showed in
Section~\ref{sec:hats}, the main cost of high availability and low
latency comes in the inability to prevent Lost Update, Write Skew, and
provide recency bounds. In this section, we attempt to understand when
these guarantees matter in a representative transactional application
based on the TPC-C benchmark~\cite{tpcc}.

@@

TPC-C consists of five transactions, capturing the operation of a
wholesale warehouse, including sales, payments, and deliveries. Two
transactions---\textit{Order-Status} and \textit{Stock-Level}---are
read-only and can be executed safely with HATs. Clients may read stale
data, but this does not violate TPC-C requirements and clients will
read their writes if they are sticky-available. Another transaction
type, \textit{Payment}, updates running balances for warehouses,
districts, and customer records and provides an audit trail. The
transaction is monotonic---increment- and append-only---so all balance
increase operations commute, and MAV allows the maintenance of
foreign-key integrity constraints (e.g., via \texttt{UPDATE/DELETE
  CASCADE}).

@@

\vspace{.5em}\noindent\textit{Integrity Constraints.} Throughout execution, TPC-C also requires the maintenance of several
integrity constraints. For example, Consistency Condition 1 (3.3.2.1)
requires that each warehouse's sales count must reflect the sum of its
subordinate sales districts. This integrity constraint spans two
tables but, given the ability to update rows in both tables atomically
via MAV, can be easily maintained. Consistency Conditions 4 through 12
(3.3.2.4-12) can similarly be satisfied by applying updates atomically
across tables. Consistency Conditions 2 and 3 (3.3.2.2-3) concern
order ID assignment and are problematic. Finally, while TPC-C is not
subject to multi-key anomalies, we note that many TPC-E isolation
tests (i.e., simultaneously modifying a product description and its
thumbnail) are also achievable using HATs.

@@

\vspace{.5em}\noindent\textit{Implementation.} Our prototype database
is a partially replicated (hash-based, partitioned) key-value backed
by LevelDB and implemented in Java using Apache Thrift. It currently
supports eventual consistency (hereafter, \texttt{eventual};
last-writer-wins RU with standard all-to-all anti-entropy between
replicas) and the efficient HAT MAV algorithm as sketched in
Section~\ref{sec:ta}. (hereafter, \texttt{MAV}). It also supports
non-HAT operation whereby all operations for a given key are routed to
a (randomly) designated \texttt{master} replica for each key
(guaranteeing single-key linearizability, as in Gilbert and Lynch's
CAP Theorem proof~\cite{gilbert-cap} and in PNUTS~\cite{pnuts}'s
\texttt{read-latest} operation; hereafter, \texttt{master}) as well as
distributed two-phase locking. Servers are durable: they synchronously
write to LevelDB before responding to client requests, while new
writes in MAV are sent to a write-ahead log, which is synchronously
flushed to disk.

@@

\vspace{.5em}\noindent\textit{Geo-replication.} We first deploy the
database prototype across an increasing number of
datacenters. Figure~\ref{fig:wan-exp}A shows that, when operating two
clusters within a single datacenter, mastering each data item results
in approximately half the throughput and double the latency of
\texttt{eventual}. This is because HAT models are able to utilize
replicas in both clusters instead of always contacting the (single)
master. \texttt{RC}---essentially \texttt{eventual} with
buffering---is almost identical to \texttt{eventual}, while
\texttt{MAV}---which incurs two writes for every client-side write
(i.e., new writes are sent to the WAL then subsequently moved into
LevelDB once stable)---achieves ~75\% of the throughput. Latency
increases linearly with the number of YCSB clients due to contention
within LevelDB.

In contrast, when the two clusters are deployed across the continental
United States (Figure~\ref{fig:wan-exp}B), the average latency of
\texttt{master} increases to $300$ms (a $278$--$4257\%$ latency
increase; average $37$ms latency per operation). For the same number
of YCSB client threads, \texttt{master} has substantially lower
throughput than the HAT configurations. Increasing the number of YCSB
clients \textit{does} increase the throughput of \texttt{master}, but
our Thrift-based server-side connection processing did not gracefully
handle more than several thousand concurrent connections. In contrast,
across two datacenters, the performance of eventual, RC, and MAV are
near-identical to a single-datacenter deployment.

When five clusters (as opposed to two, as before) are deployed across
the five EC2 datacenters with lowest communication cost
(Figure~\ref{fig:wan-exp}C), the trend continues: \texttt{master}
latency increases to nearly $800$ms per transaction. As an attempt at
reducing this overhead, we implemented and benchmarked a variant of
quorum-based replication as in Dynamo~\cite{dynamo}, where clients
sent requests to all replicas, which completed as soon as a majority
of servers responded (guaranteeing regular
semantics~\cite{herlihy-art}); this strategy (not pictured) did not
substantially improve performance due to the network topology and
because worst-case server load was unaffected. With five clusters,
\texttt{MAV}'s relative throughput decreased: every YCSB \texttt{put}
operation resulted in four \texttt{put} operations on remote replicas
and, accordingly, the cost of anti-entropy increased (e.g., each
server processed four replicas' anti-entropy as opposed to one before,
reducing the opportunity for batching and decreasing available
resources for incoming client requests). This in turn increased
garbage collection activity and, more importantly, IOPS when compared
to \texttt{eventual} and \texttt{RC}, causing \texttt{MAV} throughput
to peak at around half of \texttt{eventual}. With in-memory
persistence (i.e., no LevelDB or WAL), \texttt{MAV} throughput was
within 20\% of \texttt{eventual}.

@@

\begin{figure}[t!]
\begin{center}
\hspace{2em}\includegraphics[width=.8\columnwidth]{figs/strategylegend.pdf}\vspace{-2em}
\end{center}
\begin{center}
\includegraphics[width=\figfactor\columnwidth]{figs/finals/txnlen-thru.pdf}
\end{center}\vspace{-2.25em}
\caption{Transaction length versus throughput.}\vspace{-1em}
\label{fig:txlen}
\begin{center}
\includegraphics[width=\figfactor\columnwidth]{figs/finals/wprop-thru.pdf}
\end{center}\vspace{-2.25em}
\caption{Proportion of reads and writes versus throughput.}\vspace{-1em}
\label{fig:rprop}
\begin{center}
\includegraphics[width=\figfactor\columnwidth]{figs/finals/scaleout-thru.pdf}
%\includegraphics[width=\figfactor\columnwidth]{figs/finals/scaleout-thru-perserver.pdf}
\end{center}\vspace{-2.25em}
\caption{Scale-out of MAV, Eventual, and RC.}\vspace{-1.5em}
\label{fig:scaleout}
\end{figure}

\vspace{.5em}\noindent\textit{Transaction length.} As shown in
Figure~\ref{fig:txlen} (clusters in Virginia and Oregon), throughput
of \texttt{eventual}, \texttt{RC}, and \texttt{master} operation are
unaffected by transaction length. In contrast, \texttt{MAV} throughput
decreases linearly with increased transaction length: with $1$
operation per transaction, \texttt{MAV} throughput is within 18\% of
\texttt{eventual} ($34$ bytes overhead), and with $128$ operations per
transaction, \texttt{MAV} throughput is within $60\%$ ($1898$ bytes
overhead). This reflects our \texttt{MAV} algorithm's metadata requirements,
which are proportional to transaction length and consume IOPS and
network bandwidth.

\vspace{.5em}\noindent\textit{Read proportion.} Our default (equal)
proportion of reads and writes is fairly pessimistic: for example,
Facebook reports $99.8\%$ reads for their workload~\cite{eiger}. As
shown in Figure~\ref{fig:rprop} (clusters in Virginia and Oregon),
with all reads, \texttt{MAV} is within $4.8\%$ of \texttt{eventual};
with all writes, \texttt{MAV} is within $33\%$, and the throughput of
\texttt{eventual} decreases by $288.8\%$ compared to all reads. At
$99.8\%$ reads, \texttt{MAV} incurs a $7\%$ overhead ($5.8\%$ for in-memory
storage).

\vspace{.5em}\noindent\textit{Scale-out.} One of the key benefits of
our HAT algorithms is that they are shared-nothing, meaning they
should not compromise scalability. Figure~\ref{fig:scaleout} shows
that varying the number of servers across two clusters in Virginia and
Oregon (with $15$ YCSB clients per server) results in linear scale-out
for \texttt{eventual}, \texttt{RC}, and \texttt{MAV}. \texttt{RC} and
\texttt{eventual} scale perfectly with an approximately $5$x
throughput increase moving from $5$ to $25$ servers per cluster. For
the same configuration, \texttt{MAV} scales by $3.8$x, achieving over
$260,000$ operations per second. \texttt{MAV} performance is due to
resource contention---with a memory-backed database (instead of
LevelDB), \texttt{MAV} scales by $4.25$x (not shown)---and MAV-related
performance heterogeneity across servers (Calvin's authors report
similar heterogeneity on EC2~\cite{calvin}). Nevertheless, this linear
scalability is not typical of traditional database system
designs~\cite{gray-isolation}.

\vspace{.5em}\noindent\textit{Summary.} Our experimental prototype
confirms our earlier analytical intuitions. HAT systems can provide
useful semantics without substantial performance penalties. In
particular, our MAV algorithm can achieve throughput competitive with
eventual consistency at the expense of increased disk and network
utilization. Perhaps more importantly, all HAT algorithms circumvent
high WAN latencies inevitable with non-HAT implementations. Our results
highlight Deutsch's observation that ignoring factors such as latency
can ``cause big trouble and painful learning
experiences''~\cite{fallacies-deutsch}---in a single-site context,
paying the cost of coordination may be tenable, but, especially as
services are geo-replicated, costs increase.
